Nonparametric Goodness-of-Fit Tests for Discrete Null Distributions

### 2010-11-19 rjournal/RJtemplate.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@33 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@35 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%General idea of goodness-of-fit tests (hypothesis test about underlying distribution).
%Given an observed sequence of data, the typical question of statistical inference is to determine
%the underlying process from which it came. 
Goodness-of-fit tests are used to assess whether data are consistent
with a hypothesized null distribution.  The $\chi^2$ test is the best-known
parametric goodness-of-fit test, while the most popular nonparametric tests
are the classic test proposed by Kolmogorov and Smirnov followed closely by
several variants on Cram\'{e}r-von Mises tests.

%procedure proposed by
%Cram\'{e}r and von Mises (tests sometimes called Cram\'{e}r-von Mises-Smirnov
%tests, or simply Cram\'{e}r-von Mises tests).

%*** why "estimation procedure" above?
%*** Citations in the paragraph above?  Also check the body of paper. ***

@@

%*** Later, make sure we point out how using one of the ``old'' tests and
%hoping for the best is a problem. ***

%Given a particular null distribution, goodness-of-fit tests 
%are used in order to test whether it is likely that the data were
%generated via the null distribution.

%Why we use non-parametric tests (power to interesting shifts; robustness to underlying assumptions).
%While almost any hypothesis test can be viewed as a variant on a
%goodness-of-fit test, the term ``goodness-of-fit test'' is typically
%reserved for those tests which are nonparametric in nature. 
%That is, they do not conduct the hypothesis testing 
%through the explicit calculation of the parameters generating the data. Instead, they %generally attempt to
%calculate differences between the empirical distribution of the observed data and the overall distribution 
%of the null model. In many cases these tests are preferential since they tend to have an increased power for 
%interesting deviations from the null model, in exchange for failing to detect less interesting deviations due to 
%factors such as measurement error. By far the most popular of these nonparametric tests are due to Kolmogorov and 
%Smirnov, followed closely by several variants on an estimation procedure proposed by Cram\'{e}r and von Mises.

% Discrete versions exist; just no code
%While the original aim of nonparametric goodness of fit tests was meant for continuous null distributions, discrete versions
%have existed since the early 1970s. Unfortunately, current statistical software has largely failed to incorporate them.
%This has left the end-user to either use only parametric tests, such as Pearson's Chi-Squared test, or to incorrectly
%apply the functions meant for continuous distributions.

@@

% Most popular choice of test; explain the basic algorithm and purpose for continuous distributions (maybe a few P's).

@@

The most popular nonparametric goodness-of-fit test is the
Kolmogorov-Smirnov test.
Given the cumulative distribution
function $F_0(x)$ of the continuous null distribution
and the empirical distribution function $F_{data}(x)$ of the observed data,  the test statistic is given by
\begin{align}
D = \sup_x \left| F_0(x)- F_{data}(x) \right|.    \label{eqD}
\end{align}
The distribution of $D$ does not depend on the hypothesized
distribution, making this a computationally
attractive method. \cite{slakter65} offers a standard presentation
of the test and its
performance relative to other algorithms. 
The test statistic is easily adapted for one-sided tests.
For these, the absolute value in (\ref{eqD}) is discarded and the tests are based
on either the supremum of the remaining difference (the `greater' testing
alternative) or by replacing the supremum with a negative infimum
(the `lesser' hypothesis alternative).  Tabulated p-values have been
available for these tests since 1933. %(*** year ****).

%**** check this above, both do away with the absolute value, I think? ***

The extension of the Kolmogorov-Smirnov test to non-continuous 
null distributions is not straightforward. The formula of
the test statistic $D$ remains unchanged, but its distribution
is much more difficult to obtain; unlike the 
continuous case, its distribution depends on the null model.
It has been known since at least the 1950's that using the tables associated
with continuous null distributions results in conservative p-values
when the null distribution is discontinuous 
(see *** or *** or \citet{stephens1974}, for example).  In the early 1970's, 
\citet{Conover1972} developed the method implemented here
for computing p-values
in the case of discrete null distributions.

@@

Having obtained the test statistic, the p-value must then be calculated. 
For larger sample sizes, the null distribution can be approximated
with the null distribution from the classical Kolmogorov-Smirnov test. 
When an exact p-value is required for
smaller sample sizes, the methodology in \citet{Conover1972} is used. 
Full details of the calculations are contained in source code
of our revised function \code{ks.test()} and in Conover's original paper.

@@

%Discuss the discrete version. How to calculate W2, U2, and A2.

%*** I edited, below, make sure you check it. ****

Generalizations of the Cram\'{e}r-von Mises tests to discrete
distributions were developed in \cite{choulakian1994}. As with the
Kolmogorov-Smirnov test, the forms of the test statistics are unchanged, 
and the null distributions of the test statistics are again hypothesis-dependent. \cite{choulakian1994} does not offer finite-sample results, but rather 
shows that the asymptotic distributions of the test statistics under the null hypothesis each involve consideration of a weighted sum of independent
chi-squared variables (with the weights depending on the particular
null distribution). 
We implement the asymptotic distributions here; \cite{stephens1974}
showed that these approximations are conservative 
and asymptotically equivalent to the true null distributions.

**** Why would Stephens (1974) have this, when they were developed in 1994? ****

@@

We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \label{ineq1} \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\label{ineq2}
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most p-values, while
this upper bound is used for smaller p-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation).  Although it would be preferable to determine the use of
the bound based on values of the test statistic rather than the p-value,
the range of ``extreme'' values of the test statistic varies with the
hypothesized distribution.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1.pdf}
\end{center}
\caption{Plot of calculated p-values for given test statistics
using numerical integration
(red) compared to the conservative chi-squared bound (dashed blue). 
The null distribution is uniform on the set $\{1,2,3\}$ in this example.
The instability in the calculated p-values is a result of numerical
instabilities.}
\label{cvmissues}
\end{figure}

@@

The function \pkg{cvm.test()} is similar in design
to \pkg{ks.test()}.  Its first two
arguments specify the data and null distribution; the only extra option,
\code{type}, specifies the variant of the Cram\'{e}r-von Mises test:
\begin{itemize}
\item \code{x}: a numerical vector of data values.
\item \code{y}: an \code{ecdf} or step-function (\code{stepfun}) for specifying
the null model
\item \code{type}: the variant of the Cram\'{e}r-von Mises test; \code{W2}
is the default and most common method, \code{U2} is for cyclical data,
and \code{A2} is the Anderson-Darling alternative.
\end{itemize}
As with \code{ks.test()}, \code{cvm.test()} returns an object of class 
\code{htest}.  Although we designed \code{cmv.test()} primarily for cases with
discrete distributions \code{y}, we implemented the continuous forms of the
test for variants \code{W2} and \code{A2} (as noted in the Discussion, below).

@@

Finally, we demonstrate the Cram\'{e}r-von Mises tests with the same
simulated data. 
\begin{Schunk}
\begin{Sinput}
> library(cvm.test)
> cvm.test(x, ecdf(1:10))
\end{Sinput}
\begin{Soutput}
	Cramer-von Mises - W2

@@

*** Cyclical example? ***

\begin{Schunk}
\begin{Sinput}
> library(cvm.test)
> x <- sample(1:4, 20, replace=T)
> cvm.test(x, ecdf(1:4), type='U2')
\end{Sinput}
\begin{Soutput}
	Cramer-von Mises - U2

data:  x 
U2 = 0.075, p-value = 0.4027
alternative hypothesis: Two.sided 
\end{Soutput}
\begin{Sinput}
> y <- x%%4 + 1
> cvm.test(y, ecdf(1:4), type = 'U2')
\end{Sinput}
\begin{Soutput}
	Cramer-von Mises - U2

data:  y 
U2 = 0.075, p-value = 0.4027
alternative hypothesis: Two.sided 
\end{Soutput}
\end{Schunk}

@@

In the continuous setting, both of 
the Kolmogorov-Smirnov and the Cram\'{e}r-von Mises tests
have two-sample analogues. When data are
observed from two processes or sampled from two populations, 
the hypothesis tested is 
whether they came from the same (unspecified) distribution. An
analogous theory for discrete distributions does not exist.
This comes from the fact that the discrete null distributions
of the test statistics depend on the exact null distribution; 
therefore the two-sample case would surely have to depend
on the exact distributions used as well %(*** is this true? ***), 
which are generally not even stated in the two-sample case. 
*** I don't think I buy this. Couldn't this just be done as a
chi-squared test on a 2xp table, at least? *** Also,
see \url{http://ideas.repec.org/p/mtl/montde/2001-23.html}.

While we have implemented the two most popular variants of 
goodness-of-fit tests for discrete distributions, further 
generalizations of tests are described in the
extended study of \cite{dewev1973}.  We also provide general one-sample
Cram\'{e}r-von Mises testing for continuous distributions simply
to help complement the abilities provided by several other
R packages, although these tests are not the focus of this paper.
In particular, functions implemented in package \pkg{nortest} focus on the
composite hypothesis of normality, while
package \pkg{ADGofTest} provides the Anderson-Darling variant
of the test for continuous distributions.  Packages
\pkg{CvM2SL1Test} and \pkg{CvM2SL1Test} provide support for
two-sample Cram\'{e}r-von Mises tests with continuous distributions.
Package \pkg{cramer} offers a multivariate Cram\'{e}r test for
the two-sample problem.  Possible reference, though it isn't clear
if this particular one is really part of this literature segment?
\url{http://en.scientificcommons.org/39157548}

 *** Need to be added
to the bibliography, whatever we choose to do ***

>>>>

Goodness-of-fit tests are used to assess whether data are consistent
with a hypothesized null distribution.  The $\chi^2$ test is the best-known
parametric goodness-of-fit test, while the most popular nonparametric tests
are the classic test proposed by Kolmogorov and Smirnov followed closely by
several variants on Cram\'{e}r-von Mises tests.

@@

@@

@@

The most popular nonparametric goodness-of-fit test is the
Kolmogorov-Smirnov test.
Given the cumulative distribution
function $F_0(x)$ of the continuous null distribution
and the empirical distribution function $F_{data}(x)$ of the observed data,  the test statistic is given by
\begin{align}
D = \sup_x \left| F_0(x)- F_{data}(x) \right|.    \label{eqD}
\end{align}
The distribution of $D$ does not depend on the hypothesized
distribution, making this a computationally
attractive method. \cite{slakter65} offers a standard presentation
of the test and its
performance relative to other algorithms. 
The test statistic is easily adapted for one-sided tests.
For these, the absolute value in (\ref{eqD}) is discarded and the tests are based
on either the supremum of the remaining difference (the `greater' testing
alternative) or by replacing the supremum with a negative infimum
(the `lesser' hypothesis alternative).  Tabulated p-values have been
available for these tests since 1933.

The extension of the Kolmogorov-Smirnov test to non-continuous 
null distributions is not straightforward. The formula of
the test statistic $D$ remains unchanged, but its distribution
is much more difficult to obtain; unlike the 
continuous case, its distribution depends on the null model.
It has been known since at least the 1950's that using the tables associated
with continuous null distributions results in conservative p-values
when the null distribution is discontinuous 
(see \cite{slakter1965}, \cite{goodman1954}, and \cite{massey1951}).  
In the early 1970's, 
\citet{Conover1972} developed the method implemented here
for computing p-values
in the case of discrete null distributions.

@@

Having obtained the test statistic, the p-value must then be calculated. 
For larger sample sizes, the null distribution can be approximated
with the null distribution from the classical Kolmogorov-Smirnov test. 
When an exact p-value is required for
smaller sample sizes, the methodology in \citet{Conover1972} is used. 
Full details of the calculations are contained in source code
of our revised function \code{ks.test()} and in Conover's paper.

@@

Generalizations of the Cram\'{e}r-von Mises tests to discrete
distributions were developed in \cite{choulakian1994}. As with the
Kolmogorov-Smirnov test, the forms of the test statistics are unchanged, 
and the null distributions of the test statistics are again hypothesis-dependent. \cite{choulakian1994} does not offer finite-sample results, but rather 
shows that the asymptotic distributions of the test statistics under the null hypothesis each involve consideration of a weighted sum of independent
chi-squared variables (with the weights depending on the particular
null distribution).

@@

We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \label{ineq1} \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\label{ineq2}
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most p-values, while
this upper bound is used for smaller p-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation).  Figure \ref{cvmissues} shows this bound with the blue dashed
line; values in red exceeding the bound are a result of the numerical
instability.  Although it would be preferable to determine the use of
the bound based on values of the test statistic rather than the p-value,
the range of ``extreme'' values of the test statistic varies with the
hypothesized distribution.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1.pdf}
\end{center}
\caption{Plot of calculated p-values for given test statistics
using numerical integration
(red) compared to the conservative chi-squared bound (dashed blue). 
The null distribution is uniform on the set $\{1,2,3\}$ in this example.
The sharp variations in the calculated p-values are a result of numerical
instabilities, and the true p-values are bounded by the blue dashed curve.}
\label{cvmissues}
\end{figure}

@@

The function \code{cvm.test()} is similar in design
to \code{ks.test()}.  Its first two
arguments specify the data and null distribution; the only extra option,
\code{type}, specifies the variant of the Cram\'{e}r-von Mises test:
\begin{itemize}
\item \code{x}: a numerical vector of data values.
\item \code{y}: an \code{ecdf} or step-function (\code{stepfun}) for specifying
the null model
\item \code{type}: the variant of the Cram\'{e}r-von Mises test; \code{W2}
is the default and most common method, \code{U2} is for cyclical data,
and \code{A2} is the Anderson-Darling alternative.
\end{itemize}
As with \code{ks.test()}, \code{cvm.test()} returns an object of class 
\code{htest}.  Although we designed \code{cmv.test()} primarily for cases with
discrete distributions \code{y}, we implemented the continuous forms of the
test for variants \code{W2} and \code{A2} (as noted in the Discussion, below).

@@

We demonstrate the Cram\'{e}r-von Mises tests with the same
simulated data. 
\begin{Schunk}
\begin{Sinput}
> library(cvm.test)
> cvm.test(x, ecdf(1:10))
\end{Sinput}
\begin{Soutput}
	Cramer-von Mises - W2

@@

We conclude with a toy cyclical example showing that the test is
invariant to cyclic reordering of the support.
\begin{Schunk}
\begin{Sinput}
> library(cvm.test)
> set.seed(1)
> y <- sample(1:4, 20, replace=T)
> cvm.test(y, ecdf(1:4), type='U2')
\end{Sinput}
\begin{Soutput}
	Cramer-von Mises - U2

data:  y 
U2 = 0.0094, p-value = 0.945
alternative hypothesis: Two.sided 
\end{Soutput}
\begin{Sinput}
> z <- y%%4 + 1
> cvm.test(z, ecdf(1:4), type = 'U2')
\end{Sinput}
\begin{Soutput}
	Cramer-von Mises - U2

data:  z 
U2 = 0.0094, p-value = 0.945
alternative hypothesis: Two.sided 
\end{Soutput}
\end{Schunk}
In contrast, the Kolmogorov-Smirnov or the standard Cram\'{e}r-von Mises tests
produce different results after such a reordering. For example, the default
Cram\'{e}r-von Mises yields p-values of 
*** we have a problem, or maybe got really unlucky? ***.

@@

In the continuous setting, both of 
the Kolmogorov-Smirnov and the Cram\'{e}r-von Mises tests
have two-sample analogues. When data are
observed from two processes or sampled from two populations, 
the hypothesis tested is 
whether they came from the same (unspecified) distribution. 
This comes from the fact that the discrete null distributions
of the test statistics depend on the exact null distribution; 
therefore the two-sample case would surely have to depend
on the exact distributions used as well
which are generally not even stated in the two-sample case
\cite{newpaper}.  **** explored goodness-of-fit tests for
discrete distributions using a permutation test approach.

Further 
generalizations of goodness-of-fit tests for discrete distributions
are described in the
extended study of \cite{dewev1973}.  We also provide general one-sample
Cram\'{e}r-von Mises testing for continuous distributions simply
to help complement the abilities provided by several other
R packages, although these tests are not the focus of this paper.
In particular, functions implemented in package \pkg{nortest} 
\citep{nortest} focus on the
composite hypothesis of normality, while
package \pkg{ADGofTest} \citep{ADGofTest} provides the Anderson-Darling variant
of the test for continuous distributions.  Packages
\pkg{CvM2SL1Test} \citep{CvM2SL1Test} and \pkg{CvM2SL2Test} 
\citep{CvM2SL2Test} provide support for
two-sample Cram\'{e}r-von Mises tests with continuous distributions.
Package \pkg{cramer} \citep{cramer} offers a multivariate Cram\'{e}r test for
the two-sample problem.
